---
title: 'Useful equations for ecosystem modeling'
author: | 
  | R.Q. Thomas
  | Virginia Tech
date: "Fall 2017"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Proportional increase (or decrease)**

- Not constrained between 0 and 1

```{r}
f <- function(x,b){
  y = x*b
}
b = 0.5
curve(f(x,b=b),from=0, to=3,ylab='rate',xlab = 'driver/environmental condition')
```

**Monod or Michaelis-Menten**

- constrained between 0 and 1
- Very common
- Multiply by a maximum rate constant

```{r}
f <- function(x,ks){
  y = x/(ks + x)
}
ks = 2
curve(f(x,ks),from=0, to=20,ylab='rate adjustment',xlab = 'driver/environmental condition')
```


**Arrhenius Function (temperature sensitivity)**

- Anchored to a respiration rate at a base temperature
- Less common than Q10

```{r}
f <- function(x,Ea,R,base_rate){
  t = x
  y = base_rate*exp(-Ea/(R*t))
}
Ea = 50 #J mol-1  Activiation Energy often reported as kJ mol-1 units
R = 0.008314 #kJ mol-1 K-1 Universal Gas Constant 
base_rate = 1.92e10
curve(f(x,Ea=Ea,R = R,base_rate=base_rate),from=273, to=303,ylab='rate',xlab = 'temperature')
```

**Q10 Function  (temperature sensitivity)**

- Anchored to a respiration rate at a base temperature
- Very common

```{r}
f <- function(x,q10,base_temp,base_rate){
  y = base_rate*q10^((x-base_temp)/10)
}
q10 = 2
base_temp = 20
base_rate = 4
curve(f(x,q10=q10,base_temp = base_temp,base_rate = base_rate),from=0, to=40,ylab='rate',xlab = 'temperature')


```

**Exponential increase**

- not constrained between 0 and 1

```{r}
f <- function(x,k){
  y = exp(k*x)
}
k = 1.1
curve(f(x,k),from=0, to=4,ylab='rate adjustment',xlab = 'driver/environmental condition')

```

**Exponential decrease**

- constrained between 0 and 1

```{r}
f <- function(x,k){
  y = exp(k*x)
}
k = -2
curve(f(x,k),from=0, to=4,ylab='rate adjustment',xlab = 'driver/environmental condition')

```

**Optimal function without symetry**

- constrained between 0 and 1
- common for modeling photosynthesis - temperature relationships

```{r}
f <- function(x,Tmin,Topt,Tmax){
  y = ((x - Tmin) / (Topt - Tmin)) *((Tmax - x) / (Tmax - Topt)) ^((Tmax - Topt) / (Topt - Tmin))
}

Tmin = 0
Topt = 10
Tmax = 17
curve(f(x,Tmin=Tmin,Topt=Topt,Tmax=Tmax),from=0, to=20,ylab='rate adjustment',xlab = 'driver/environmental condition')
```

**Optimal function with symetry**

- Can have negative inputs
- Bounded between 0 and 1

```{r}
f <- function(x,a,b){
  y = exp(-(1/(2*b))*(x-a)^2)
}

curve(f(x,a = 1, b = 2),from=-6, to=6,ylab='rate limitation',xlab = 'driver/environmental condition')
```


**Non-linear decrease**

- constrained between 0 and 1
- Useful for simulating senstivity to soil water avialablity

```{r}
f <- function(x,a,b){
  y = (1/(1 + (x/a)^b))
}

a = 10
b = 5
curve(f(x,a=a,b=b),from=0, to=20,ylab='rate adjustment',xlab = 'driver/environmental condition')
a = 15
b = 5
curve(f(x,a=a,b=b),from=0, to=20,ylab='rate adjustment',xlab = 'driver/environmental condition',add=TRUE,lty='dashed')
```

**Non-linear decrease V2**

- constrained between 0 and 1
- Useful for simulating senstivity to soil water avialablity

```{r}
f <- function(x,a,b){
  y = 1 / (1 + exp(a*(x - b)))
}

a = 0.3
b = 5
curve(f(x,a=a,b=b),from=0, to=20,ylab='rate adjustment',xlab = 'driver/environmental condition')
segments(b,0,b,0.5,lty='dashed')
segments(0,0.5,b,0.5,lty='dashed')
```

**Non-linear increase** 

- requires relative environmental conditions (i.e., proportion of max soil water)
- Constrained between 0 and 1

```{r}
f <- function(x,a,b){
  y = (1/(1 + ((1-x)/a)^b))
}

a = 0.5
b = 5
curve(f(x,a=a,b=b),from=0, to=1,ylab='rate adjustment',xlab = 'driver/environmental condition')
a = 0.75
b = 5
curve(f(x,a=a,b=b),from=0, to=1,ylab='rate adjustment',xlab = 'relative environmental condition',add=TRUE,lty='dashed')
```

**Light extinction (variation of the exponential decay)**

- Represents at cumulative total
- Bounded between 0 and 1
- Beer's Law

```{r}
f <- function(x,k){
  y = (1.0 - (exp(-k * x)))
}

k = 0.5
curve(f(x,k = 0.5),from=0, to=10,ylab='proportion of light avaialable absorbed by layer',xlab = 'depth of lake or leaf area')

```

**Beta function**

- Not bounded between 0 and 1
- A value of 1 at a base value for the driver/environmental condition
- Example is for infuence of atmospheric CO2 on photosynthesis

```{r}
f <- function(x,beta,base){
  (1 + beta*log(x/base) )
}
base = 350
beta = 1.4
curve(f(x,beta = beta, base = base),from=200, to=700,ylab='rate limitation',xlab = 'driver/environmental condition')
segments(base,0,base,1.0,lty='dashed')
segments(200,1.0,base,1.0,lty='dashed')
```

**Lognormal function**

- Bounded between 0 and 1
- Requires input values greater than 0

```{r}
f <- function(x,a,b){
  y = exp(-(1/2*b)*(log(x)-a)^2)
}
curve(f(x,a = 1, b = 2),from=0, to=10,ylab='rate limitation',xlab = 'driver/environmental condition')
```

**Power law**

- Not bounded between 0 and 1

```{r}
f <- function(x,a,k){
  y = a*x^k
}
curve(f(x,a = 1, k = 0.5),from=0, to=10,ylab='rate limitation',xlab = 'driver/environmental condition')
```
